Nail: A practical interface generator for data formats

### 2014-05-01 paper/osdi14/design.tex
pass over 3.4

\subsection{Input streams and Transformations.}
\label{s:transforms}
Traditional parsers handle input one symbol at a time, beginning to end.
However, real-world formats often require non-linear parsing. Offset fields require a parser to move
to a different position in the input, possibly backwards. Size fields require the parser to stop
processing before the end of input has been reached and perhaps continue executing a parent parser.
Other cases, such as compressed data require more complicated processing on parts of the input
before it can be handled.

Nail introduces two concepts to handle this, \emph{streams} and \emph{transformations}. 
Streams represent a sequence of bytes that contain some external format. The parsers and generators
Nail generates always operate on an implicit stream named \cc{\$current} that they process front to
back, reading input or appending output.

Grammars can use the \cc{apply} combinator to parse or generate external data on a different stream,
inserting the result in the data model. Streams can be defined in two ways.

First, a rule can require streams as arguments, which can be passed in by the program that invokes
the generated parser.Every rule needs toe passed at least the implicit current stream.

Second, the programmer can create temporary streams in the grammars with \emph{transformations},
which are two arbitrary functions called during parsing and output generation.
The parsing function  takes any number of stream arguments and dependent field values
and produces any number of temporary streams. This function may reposition and read from the
input streams and read the values of dependent fields, but not change their contents and values. 
The generating function has to be an inverse of the parsing function. It takes the same number of temporary streams
the parsing function produces and and writes the same number of streams and dependent field
values the parsing function consumes.

Typically, the top level of most grammars is a rule that only takes a single stream, which may then
be broken up by various transformations and passed to sub-rules, which eventually parse various linear
fragment streams. Upon parsing, these fragment streams are generated and then combined by the
transforms.

The programmer has to guarantee memory safety and unambiguity of the stream transformations. To
reduce both programmer effort and the risk of unsafe operations, we provide implementations of
transformations for many common features, such as checksums and size and offset fields. Furthermore,
we provide library functions that can be used to safely operate on streams, such as splitting and
concatenation. Nail implements streams as iterators, so they can share underlying buffers and can be
efficiently duplicated and split.

Transformations can handle a wide variety of patterns in data formats, including the following: 
\paragraph{Offsets.}
We provide a transformation for handling offset fields that is invoked as follows:
\cc{\$fragment transform offset\_u32(\$file\_stream, \@offset)}. 
They are implemented by two functions, as shown in Figure~\ref{fig:xform-sig}.

@@

\paragraph{Compressed data.}
Encoded, compressed or encrypted data can be handled transparently by writing a custom
transformation that transforms a coded stream into one that can be parsed by a Nail grammar and vice
versa. Usually, this involves calling an existing library.

\paragraph{Checksums.}
Checksums can be verified and computed in a transformation that takes a stream and a dependent
field. In some cases, a checksum is calculated over a buffer that contains the checksum itself, with
the checksum being set to some particular value. Because the functions implementing a transformation
are passed a pointer to any dependent fields, the checksum function can set the checksums initial
value before calculating the checksum over the entire buffer, including the checksum.

\noindent A real-world example with many different transforms is explained in Section~\ref{s:eval-format-zip}.

>>>>

\subsection{Input streams and transformations}
\label{s:transforms}

Traditional parsers handle input one symbol at a time, from beginning to end.
However, real-world formats often require non-linear parsing. Offset fields require a parser to move
to a different position in the input, possibly backwards. Size fields require the parser to stop
processing before the end of input has been reached, and perhaps resume executing a parent parser.
Other cases, such as compressed data, require more complicated processing on parts of the input
before it can be handled.

Nail introduces two concepts to handle this, \emph{streams} and \emph{transformations}. 
Streams represent a sequence of bytes that contain some external format. The parsers and generators
that Nail generates always operate on an implicit stream named \cc{\$current} that they process front to
back, reading input or appending output.
Grammars can use the \cc{apply} combinator to parse or generate external data on a different stream,
inserting the result in the data model.

Streams can be defined in two ways.
First, a rule can require streams as arguments, which can be passed in by the program that invokes
the generated parser. Every rule needs to be passed at least the implicit current stream.

Second, the programmer can create temporary streams in the grammar with \emph{transformations},
which are two arbitrary functions called during parsing and output generation respectively.
The parsing function  takes any number of stream arguments and dependent field values,
and produces any number of temporary streams. This function may reposition and read from the
input streams and read the values of dependent fields, but not change their contents and values. 
The generating function has to be an inverse of the parsing function. It takes the same number of
temporary streams that the parsing function produces, and writes the same number of streams and
dependent field values that the parsing function consumes.

Typically, the top level of most grammars is a rule that takes only a single stream, which may then
be broken up by various transformations and passed to sub-rules, which eventually parse various linear
fragment streams. Upon parsing, these fragment streams are generated and then combined by the
transforms.

The programmer has to guarantee memory safety and unambiguity of the stream transformations. To
reduce both programmer effort and the risk of unsafe operations, we provide implementations of
transformations for many common features, such as checksums, size, and offset fields. Furthermore,
we provide library functions that can be used to safely operate on streams, such as splitting and
concatenation. Nail implements streams as iterators, so they can share underlying buffers and can be
efficiently duplicated and split.

Transformations can handle a wide variety of patterns in data formats, including the following: 
\paragraph{Offsets.}
We provide a transformation for handling offset fields, which is invoked as follows:
\cc{\$fragment transform offset\_u32(\$file\_stream, \@offset)}. 
This transformation corresponds to two functions for parsing and
generation, as shown in Figure~\ref{fig:xform-sig}.

@@

\paragraph{Compressed data.}
Encoded, compressed, or encrypted data can be handled transparently by writing a custom
transformation that transforms a coded stream into one that can be parsed by a Nail grammar and vice
versa. Usually, this involves calling an existing library.

\paragraph{Checksums.}
Checksums can be verified and computed in a transformation that takes a stream and a dependent
field. In some cases, a checksum is calculated over a buffer that contains the checksum itself, with
the checksum being set to some particular value. Because the functions implementing a transformation
are passed a pointer to any dependent fields, the checksum function can set the checksum's initial
value before calculating the checksum over the entire buffer, including the checksum.

\paragraph{}
A real-world example with many different transforms, used to support
the Zip file format, is described in \S\ref{s:eval-format-zip}.
